Evolution of Processes in Delivering Reliable Software in the European Scientific Arena

### 2018-06-29 paper-ieee-trans.tex
Add Cris comment

From the advent of Grid technology -- as the new paradigm of distributed
computing -- to the current days of Cloud computing models, the continuous need
of new tools and services to match the scientific community requirements has been
addressed in Europe by various projects dedicated to software development
and e--Infrastructure creation, operation and management.
This work examines the evolution of the software engineering methodologies used in past and
present European Commission--funded projects to illustrate how the research on software
reliability has progressed in Europe over the last 15 years following the foundation of the
first project in 2001. In particular, we will summarize the techniques and procedures
applied to deliver quality software, highlighting the main challenges and barriers confronted
at the time, and how they were partially or totally overcome. The knowledge base established
throughout the years, sustained by the advances in the area of software engineering,
definitely constituted a significant breakthrough in the reliability of software delivered in the
European research arena. Recent software development projects funded by the European
Commission are good evidence of such insights, where the enforcement of Quality Assurance
procedures is present since the very early stages of the software development process.

@@

A significant number of innovation projects and technology transfer
activities -- later discussed in Section~\ref{sec:ev} -- from the European Commission
(EC)-funded 5th, 6th and 7th Framework Programmes, and recently from the Horizon 2020 programme
\cite{h2020} had as their main goal to develop software solutions that provide a seamless and
transparent access by researchers to distributed computing and data resources.

@@

In this section, we detail the level of knowledge requested in delivering reliable software in order 
to understand the effort employed in the EC-funded software projects over time. We divide literature 
into four different domains that intersect this work: 1) software requirements, 2) software coding,
3) metrics and measurements, 4) testing.

\emph{Software requirements.} They are defined at various stages of a software development process
as a specification of what should be implemented. According to IEEE \cite{radatz}, requirement
analysis is a process for studying user needs to arrive at a definition of system, hardware or software
requirements. Characteristics of requirements that can assess software reliability are consistency and
correctness \cite{sommerville}: the former aims at providing predictable, maintainable and reliable
results, and the latter usually refers to the satisfaction of certain goals.

\emph{Software coding.} According to IEEE \cite{radatz}, coding is the process of expressing a computer
program in a programming language. To meet specific requirements, a program can be developed by using
high-level or low-level programming languages that envelope different characteristics and may affect the
result and quality of software. Furthermore, software is generally written by geographically distributed
teams whose developers can change over time, therefore it is extremely important for source code to be
understandable \cite{tashtoush}. There are various instruments that can be adopted during the software
development process to easily adapt software \cite{buse, baggen}: programming features that influence the
code readability and maintainability and standardized code that promotes the comparability of results.

@@

\emph{Testing.} According to IEEE \cite{radatz}, testing is a process of operating a system or component
under specified conditions, observing or recording the results, and making an evaluation of some aspect of
system or component. Testing is an important factor for software quality and reliability improvement. This
activity is quite complicated due to the variety of programming languages, operating systems and hardware
platforms that evolve over time. In the current literature, there are many types of software
testing methods \cite{myers} that aim at finding errors, bugs and faults during the software development
process, possibly in the early stages. Over time, testing metrics have been defined to assess software
quality \cite{athanasiou}: code coverage is the most frequently used metric for test software reliability
and there exist many tools for dynamic code coverage estimation according to the programming language
\cite{horgan}. A dedicated effort have to be taken into account in order to integrate testing solutions
during the software development process.

@@

The Agile manifesto \cite{agile-manifesto} was by then an emerging methodology, not yet
considered by the project, and as such the project's
development design suffered from the lack of the methods, tools, techniques and best
practices emerging from this new discipline of software engineering
\cite{agile}. Nevertheless, the project focused on the experiences and procedures
used by diverse collaborative open source software projects, such as Linux and Apache, trying to
achieve a higher maturity level \cite{cmm}.

@@

The three phases of Enabling Grids for E--sciencE ({\sl EGEE}, Apr 2004 -- Apr 2010)
\cite{cordis:egee,cordis:egee2,cordis:egee3} projects brought together
scientists and engineers from over 240 institutions in 45 countries aiming to
provide a seamless Grid infrastructure for e--Science. {\sl EGEE--II} and {\sl EGEE--III}
featured the internationalization (outside Europe) of the project, embracing worldwide research
institutions and user communities~\cite{egee}. The software to sustain the increasing
requirements coming from the diverse scientific communities needed developing a
rich set of new services while maintaining a sustainable infrastructure for
Grid computing, eventually used by more than 15 thousand researchers and deployed in
over 250 institutions.

@@

The E--Infrastructure for Testing, Integration and Configuration of Software~\cite{etics}
({\sl ETICS}, Jan 2006 -- Feb 2010) project aimed
at addressing the challenges in producing quality software in distributed,
collaborative projects such as {\sl EGEE} and its {\sl gLite} middleware. The framework
integrated different technologies and tools in order to provide automated configuration,
build and testing capabilities, as well as auto--generated documentation and
software metrics gathering such as Source Lines Of Code (SLOC), complexity and
number of defects/bugs \cite{etics}. The {\sl ETICS} framework was the first automated
service for delivering quality software products in distributed environments like
the Grids.

@@

For each software characteristic, a set of associated
metrics and Key Performance Indicators (KPIs) were identified and defined in
detail in the {\sl EMI} Metrics Specification \cite{emi-quality-model}. The project
leveraged the {\sl ETICS} service for the development, continuous integration and release management,
as well as for metric tracking, making queries on the collected data to display
them through the chart generation framework.

@@

The {\sl EGI--InSPIRE} (Integrated Sustainable Pan--European Infrastructure for
Researchers in Europe, May 2010 -- May 2014) project \cite{cordis:egi-inspire}
was the continuation of the {\sl EGEE--III} project, with the objective of establishing and
maintaining a sustainable European Grid Infrastructure, composed by a federation of National
Grid Initiatives and interoperable with other Grids worldwide.
This goal would be accomplished through the
development and maintenance of various operational tools, such as the Operations Portal \cite{egi-ops},
the {\sl EGI} Helpdesk \cite{ggus}, or the Grid configuration database (GOCDB) \cite{gocdb}, and the
management of the software provisioning process \cite{mario}, dealing with the validation and distribution of the software to
the production infrastructures. As a result, this process led to a production--ready
Grid computing middleware distribution named {\sl UMD} (Unified Middleware Distribution). At that time the {\sl UMD}
was a stack of about 250 software components from several technology providers, such
as {\sl EMI} and {\sl Globus}. In order to be distributed through UMD
repositories, the software had to be compliant with EGI's Quality Criteria (QC) \cite{egi-qc}. The
QC defined a set of requirements in different areas: documentation, deployment, security, information
model and operations.
{\sl UMD} is still being used and
deployed in the European scientific e--Infrastructures under the follow--up
project {\sl EGI--Engage} (Engaging the EGI Community towards an Open Science
Commons) \cite{cordis:egi-engage}. This distribution is currently complemented
by a Cloud--specific one called {\sl CMD} (Cloud Middleware Distribution).

@@

The SQA criteria are a set of conventions and recommendations that pave the way for
the adequate development, timely delivery and reliable operation of the produced software components.
It emphasizes the quality requirements and best practices to be applied at the
development phase, such as style and testing compliance or human code reviews, to protect the
production source code versions. Furthermore, the criteria promote the adoption of the software
by providing minimum requirements of documentation content and stressing the usage of automated
solutions for software deployment. The SQA criteria are continuously evolving \cite{sqa-baseline}, aiming at serving
as a reference for future software development projects within the European research ecosystem.

@@

\item \textit{Unit and Functional testing}.
Changes involving the addition of new functionalities, are required to be tested. \textit{Regression
testing} in this context is accomplished by enforcing the definition and periodic execution
of the tests, covering past fixed issues, to ensure that they are not reintroduced during the development
activities. \textit{Unit testing} completes the source code testing evaluation by focusing on the code's
internal design. The SQA criteria of {\sl INDIGO--DataCloud} set a recommended threshold of 70\% coverage
for unit testing. In Fig.~\ref{fig:fig_unittest}, unit testing coverage is compared for each software
component between the two major releases, denoting an incremental trend over the course of the project:
code coverage increased from an average value of 50.32\% to 63.34\%, outlined by the respective dates of
the project's first (INDIGO--1) and second (INDIGO--2) releases. By INDIGO--2, 53\% of the software components
(17 out of 32) were over the threshold of the aforementioned project's SQA code coverage recommendation
(70\%), while 75\% (24 out of 32) of the product stack exceeded 50\% coverage. 
The indiviual records followed also a growing trend (only 6 out of 32 components had lower values in 
INDIGO--2). It is worth noting that Fig.~\ref{fig:fig_unittest} not only includes new software developments 
implemented from scratch within the project. Software tools and libraries from 
external open--source projects -- such as {\sl tosca-parser} or {\sl heat-translator} --, contributed upstream
by the project, and well--established products, involved in {\sl INDIGO--DataCloud}
but not contributing with the 100\% of their codebases -- such as {\sl dCache} --, are also considered in the
analysis. However the values given apply to the entire codebase, even for the latter case of products. These 
products challenged the application of the SQA policies described in this section, by two means:
\begin{itemize}
    \item Previously developed code was not refactored since it was not \textit{owned} by the project.
    \item An agreement on prevailing SQA policies for such cases where quality practices were already in place.
\end{itemize}

@@

\item \textit{Automated Deployment}.
To lower the barriers of software adoption, the SQA criteria require the automated
deployment of the products delivered as part of the catalogue. Automation in this context is
tackled using a configuration management tool. These tools allow the adoption of
\textit{Infrastructure as Code} (IaC), managing the component's deployment through declarative
definitions. The definition files appear as an additional source of documentation
-- self--documenting code -- as they sequentially guide the component's deployment process on multiple
platforms. {\sl INDIGO--DataCloud} project contributed to open--source IaC tools such
as Ansible
and Puppet\footnote{\url{https://forge.puppet.com/indigodc}}. A representative example of
such contributions are the \textit{50 roles} developed from scratch and currently hosted in
the Ansible Galaxy portal\footnote{\url{https://galaxy.ansible.com/indigo-dc/}}. Fig.~\ref{fig:fig_confman}
shows the number of products that offer an automated means for deployment.
It shows an increase in the adoption of such tools leading to the {\sl INDIGO--1} release
(dark blue points and trend line), as well as to the {\sl INDIGO--2} release
(light cream points and trend line). The rate of adoption is lower during the weeks
before the second release because a significant fraction of the products had already adopted it
previously to the first release.
\end{itemize}

@@

\textit{Development metrics} are obtained programmatically from several sources, namely GitHub
API\footnote{\url{https://developer.github.com/v3/}.}
and Jenkins\footnote{\url{https://jenkins.io/}} service, and graphically displayed as GitHub
pages using the GrimoireLab framework\footnote{\url{http://grimoirelab.github.io/}}.
Per--component weekly reports, including the SQA requirement
fulfilment, are issued and individually discussed through the different communication channels
with the development teams. Issue tracking metrics and KPIs are an essential piece of information
of both feature addition and defect solving, useful to detect and fix misbehaviours while in the
development phase.

@@

As any new change, involving a new feature or fix, passes through the automated
SQA machinery, the software is always ready to be released. As a result, the often compromised harmonization
between the continouous release of new features -- pushed by the development teams -- and the maintenance of
stable production systems -- demanded by users and resource providers -- can be guaranteed. The DevOps culture
theorizes and provides practical solutions that aim at unifying both software development (dev) and software
operations (ops) teams, by emphasizing the SQA techniques to avoid infrastructure disruption whenever new
developments are deployed into production systems. 
The incremental adoption of DevOps approaches throughout the project's lifetime was a key player towards
software reliability. Starting with a CI scenario at the first stages of the project, the software delivered
was eventually validated and distributed using CD pipelines.

@@

Fig.~\ref{fig:fig_jenkins_CI_builds} shows the evolution, throughout the
{\sl INDIGO--DataCloud}'s first and second releases, of the required test types builds since the
implementation of the CI infrastructure. Automated functional testing coverage were not
available for all the software stack, thus the associated number of builds are fewer. Towards the
end of the project, the software development activity slowed the pace but not completely ceased
as the CI infrastructure is still maintained and used after the project's end of life.

\subsubsection{Continuous Delivery}
As DevOps suggests, frequent releases positively affect the reliability of the software as
they reduce the time in which a given defect is exposed or has an impact, allowing the 
development teams to act promptly based on the regular feedback \cite{chen2015}. The software updates of
{\sl INDIGO--DataCloud} products, that have been taking place since the second major release, are passing through
a Continuous Delivery (CD) pipeline that adds the packaging of the software right after the
successful execution of the CI pipeline previously described. Consequently, CI--validated
software components are automatically delivered in online repositories. Nonetheless,
expert supervision to validate the results is required.

@@

Two applications from "Definition of Support to Research Communities" work package,
DisVis \cite{disvis} and PowerFit \cite{powerfit}, were
integrated into a similar CI/CD pipeline described in section \ref{sec:devops}.
Fig.~\ref{fig:fig_disvis} shows the pipeline for the Disvis application.

@@

The novelty introduced in the pipeline above is the validation of the application.
Once the application is packaged as a Docker image, and subsequently uploaded
to the Dockerhub repository, it is instantiated in a new container to be validated.
The application is then executed and the results compared with a set of reference outputs.
This pipeline implementation goes a step forward by testing the application
execution for the latest available Docker image in the catalogue.

@@

Two pilot infrastructures are at the disposal of developers and scientific
communities involved in the project. The aim of these testbeds is to test the
level of integration between the components involved in the {\sl INDIGO--DataCloud}
solutions and use cases validation, by deploying and executing the applications
with the last stable version of the software in environments as similar as
possible to production ones. A map of the Pilot Preview
infrastructure is depicted in Fig.~\ref{fig:fig_pilotpreview}. It shows the
resource providers and the components or services they deployed and supported.

@@

% Comment
%%P8 C1 L26: What is the demonstration that reliability has been improved significantly with the addition of stricter methologies? Is there some quantitative evidence of uptime, or bug reports per user, or similar? Or some evidence of speed of defect detection & fixing, which you cite as greatly improved? I am very sympathetic to the idea that systematic development techniques can help improve software quality (without getting into circular definitions, where "quality" means that methodologies or standards have been applied!) but there will also be a point at which the methodology gets too heavy and compromises the "agility" of the development team: how can you identify which processes are helping and which are not?
Reliability of the software has been substantially improved by applying the SQA
criteria in the development stage. Consequently, each change in the source code
that is meant to be integrated in the production version, is validated according
to the SQA testing requirements. Thus, defects are detected and corrected early in the
software lifecycle, reducing the cost of fixing errors at later stages. Automation is
key in this process, allowing the implementation of a CI scenario that enforces
the SQA criteria compliance. The application of more advanced DevOps practices
resulted in the implementation of CD pipelines to extend automation to the
delivery of packages. Thereby, safe and frequent software releases are possible, reducing
the impact on production systems. In this regard, a pre--production validation,
through the preview testbeds or resource center’s early adoption -- handled by the
Staged--rollout process -- alleviates the likelihood of discovering uncovered
issues at production.

The software lifecycle demands as well a continuous measurement and analysis
process, that allows timely reactions to discover and fix defects throughout the
development, release and maintenance stages. Conscientious metric
tracking and response sits at the core of an appropriate software maturity, as
ultimate reliability can only be achieved by enforcing and empowering the SQA
criteria through each stage of the software lifecycle.

@@

\bibitem{datagrid}
L. Momtahan and A. Martin,
``e--Science Experiences: Software Engineering Practice and the EU DataGrid,''
in \emph{9th Asia--Pacific Softw. Eng. Conf.},
pp. 269--275, Dec. 2002.

@@

\bibitem{cordis:egi-inspire}
I. C. Plasencia,
``EGI.eu the European grid initiative,''
in \emph{4th Iberian Grid Infra. Conf. Proc.},
pp. 5--15, 2010.

@@

\bibitem{rafi2012}
D. M. Rafi \emph{el al.},
``Benefits and limitations of automated software testing: Systematic literature review and practitioner survey,''
in \emph{Proc, 7th Int. Workshop Automation Softw. Test},
36--42, 2012.

>>>>

From the advent of Grid technology -- as the new paradigm of distributed
computing -- to the current days of Cloud computing models, the continuous need
of new tools and services to match the scientific community requirements has been
addressed in Europe by various projects dedicated to software development
and e--Infrastructure creation, operation and management.
This work examines the evolution of the software engineering processes used in past and
present European Commission--funded projects to illustrate how the research on software
reliability has progressed in Europe over the last 15 years following the foundation of the
first project in 2001. In particular, we will summarize the techniques and procedures
applied to deliver quality software, highlighting the main challenges and barriers confronted
at the time, and how they were partially or totally overcome. The knowledge base established
throughout the years, sustained by the advances in the area of software engineering,
definitely constituted a significant breakthrough in the reliability of software delivered in the
European research arena. Recent software development projects funded by the European
Commission are good evidence of such insights, where the enforcement of Quality Assurance
procedures is present since the very early stages of the software development process.

@@

A significant number of innovation projects and technology transfer
activities -- later discussed in Section~\ref{sec:ev} -- 
were funded by the 5th, 6th and 7th Framework Programmes of the European Commission (EC) 
and recently by the Horizon 2020 programme \cite{h2020}. Their main goal is to develop 
software solutions that provide a seamless and transparent access by researchers to 
distributed computing and data resources.

@@

In this section, we detail the level of knowledge requested in delivering reliable software in order 
to understand the effort employed in the EC-funded software projects over time. We divide literature 
into four different domains that intersect this work: 1) software requirements analysis, 2) software
coding,
3) metrics and measurements, 4) testing.

\emph{Software requirements analysis.} They are defined at various stages of a software development process
as a specification of what should be implemented. According to IEEE \cite{radatz}, requirement
analysis is a process for studying user needs to arrive at a definition of system, hardware or software
requirements. Characteristics of requirements that can assess software reliability are consistency and
correctness \cite{sommerville}: the former aims at providing predictable, maintainable and reliable
results, and the latter usually refers to the satisfaction of certain goals.

\emph{Software coding.} According to IEEE \cite{radatz}, coding is the process of expressing a computer
program in a programming language. To meet specific requirements, a program can be developed by using
high-level or low-level programming languages that envelope different characteristics and may affect the
result and quality of software. Furthermore, software is generally written by geographically distributed
teams whose developers can change over time, therefore it is extremely important for source code to be
understandable \cite{tashtoush}. There are various instruments that can be adopted during the software
development process to easily adapt software: programming features that influence the
code readability and maintainability \cite{buse} and standardized code that promotes the comparability of results \cite{baggen}.

@@

\emph{Testing.} According to IEEE \cite{radatz}, testing is a process of operating a system or component
under specified conditions, observing or recording the results, and making an evaluation of some aspect of
the system or component. Testing is an important factor for software quality and reliability improvement. This
activity is quite complicated due to the variety of programming languages, operating systems and hardware
platforms that evolve over time. In the current literature, there are many types of software
testing methods \cite{myers} that aim at finding errors, bugs and faults during the software development
process, possibly in the early stages. Over time, testing metrics have been defined to assess software
quality \cite{athanasiou}: code coverage is the most frequently used metric for testing software reliability
and there exist many tools for dynamic code coverage estimation according to the programming language
\cite{horgan}. A dedicated effort has to be taken into account in order to integrate testing solutions
during the software development process.

@@

The Agile manifesto \cite{agile-manifesto} was by then an emerging methodology, not yet
considered by the project. As such, the project's
development design suffered from the lack of the methods, tools, techniques and best
practices emerging from this new discipline of software engineering
\cite{agile}. Nevertheless, the project focused on the experiences and procedures
used by diverse collaborative open source software projects, such as Linux and Apache, trying to
achieve a higher maturity level \cite{cmm}.

@@

The three phases of Enabling Grids for E--sciencE ({\sl EGEE}, Apr 2004 -- Apr 2010)
\cite{cordis:egee,cordis:egee2,cordis:egee3} projects brought together
scientists and engineers from over 240 institutions in 45 countries aiming to
provide a seamless Grid infrastructure for e--Science. {\sl EGEE--II} and {\sl EGEE--III}
featured the internationalization (outside Europe) of the project, embracing worldwide research
institutions and user communities~\cite{egee}. The software to sustain the increasing
requirements coming from the diverse scientific communities needed developing a
rich set of new services, while maintaining a sustainable infrastructure for
Grid computing. This e-Infrastructure was eventually used by more than 15 thousand researchers
and deployed in over 250 institutions.

@@

The E--Infrastructure for Testing, Integration and Configuration of Software~\cite{etics}
({\sl ETICS}, Jan 2006 -- Feb 2010) project aimed
at addressing the challenges in producing quality software in distributed,
collaborative projects such as {\sl EGEE} and its {\sl gLite} middleware. The framework
integrated different technologies and tools in order to provide automated configuration,
build and testing capabilities, as well as auto--generated documentation and
software metrics gathering -- such as Source Lines Of Code (SLOC), complexity and
number of defects/bugs \cite{etics} --. The {\sl ETICS} framework was the first automated
service for delivering quality software products in distributed environments like
the Grids.

@@

For each software characteristic, a set of associated
metrics and Key Performance Indicators (KPIs) were identified and defined in
detail in the {\sl EMI} Metrics Specification \cite{emi-quality-model}. The project
leveraged the {\sl ETICS} service for the development, continuous integration and release management,
as well as for metric tracking, making queries on the collected data to display
them through a chart generation framework.

@@

The {\sl EGI--InSPIRE} (Integrated Sustainable Pan--European Infrastructure for
Researchers in Europe, May 2010 -- May 2014) project \cite{cordis:egi-inspire}
was the continuation of the {\sl EGEE--III} project, with the objective of establishing and
maintaining a sustainable European Grid Infrastructure, composed by a federation of National
Grid Initiatives and interoperable with other Grids worldwide.
This goal would be accomplished through the
development and maintenance of various operational tools -- such as the Operations Portal \cite{egi-ops},
the {\sl EGI} Helpdesk \cite{ggus}, or the Grid configuration database (GOCDB) \cite{gocdb} -- and the
management of the software provisioning process \cite{mario}, dealing with the validation and distribution of the software to
the production infrastructures. As a result, this process led to a production--ready
Grid computing middleware distribution named {\sl UMD} (Unified Middleware Distribution). At that time the {\sl UMD}
was a stack of about 250 software components from several technology providers, such
as {\sl EMI} and {\sl Globus}. In order to be distributed through UMD
repositories, the software had to be compliant with EGI's Quality Criteria (QC) \cite{egi-qc}. The
QC defined a set of requirements in different areas: documentation, deployment, security, information
model and operations.
{\sl UMD} is still being used and
deployed in the European scientific e--Infrastructures under the follow--up
project {\sl EGI--Engage} (Engaging the EGI Community towards an Open Science
Commons) \cite{cordis:egi-engage}. This distribution is currently complemented
by a Cloud--specific one called {\sl CMD} (Cloud Middleware Distribution).

@@

The SQA criteria are a set of conventions and recommendations that pave the way for
the adequate development, timely delivery and reliable operation of the produced software components.
They emphasize the quality requirements and best practices to be applied at the
development phase, such as style and testing compliance or human code reviews, to protect the
production source code versions. Furthermore, the criteria promote the adoption of the software
by providing minimum requirements of documentation content and stressing the usage of automated
solutions for software deployment. The SQA criteria are continuously evolving \cite{sqa-baseline}, aiming at serving
as a reference for future software development projects within the European research ecosystem.

@@

\item \textit{Unit and Functional testing}.
Changes involving the addition of new functionalities are required to be tested. \textit{Regression
testing} in this context is accomplished by enforcing the definition and periodic execution
of the tests, covering past fixed issues, to ensure that they are not reintroduced during the development
activities. \textit{Unit testing} completes the source code testing evaluation by focusing on the code's
internal design. The SQA criteria of {\sl INDIGO--DataCloud} set a recommended threshold of 70\% coverage
for unit testing. In Fig.~\ref{fig:fig_unittest}, unit testing coverage is compared for each software
component between the two major releases, denoting an incremental trend over the course of the project:
code coverage increased from an average value of 50.32\% to 63.34\%, outlined by the respective dates of
the project's first (INDIGO--1) and second (INDIGO--2) releases. By INDIGO--2, 53\% of the software components
(17 out of 32) were over the threshold of the aforementioned project's SQA code coverage recommendation
(70\%), while 75\% (24 out of 32) of the product stack exceeded 50\% coverage. 
The individual records followed also a growing trend (only 6 out of 32 components had lower values in 
INDIGO--2). It is worth noting that Fig.~\ref{fig:fig_unittest} not only includes new software developments 
implemented from scratch within the project. Software tools and libraries from 
external open--source projects -- such as {\sl tosca-parser} or {\sl heat-translator} --, contributed upstream
by the project, and well--established products, involved in {\sl INDIGO--DataCloud}
but not contributing with the 100\% of their codebases -- such as {\sl dCache} --, are also considered in the
analysis. However the values given apply to the entire codebase, even for the latter case of products. These 
products challenged the application of the SQA policies described in this section, by two means:
\begin{itemize}
    \item Previously developed code was not refactored since it was not \textit{owned} by the project.
    \item An agreement on prevailing SQA policies for such cases where quality practices were already in place.
\end{itemize}

@@

\item \textit{Automated Deployment}.
To lower the barriers of software adoption, the SQA criteria require the automated
deployment of the products delivered as part of the catalogue. Automation in this context is
tackled using configuration management tools. These tools allow the adoption of
\textit{Infrastructure as Code} (IaC) practice, managing the component's deployment through declarative
definitions. The definition files appear as an additional source of documentation
-- self--documenting code -- as they sequentially guide the component's deployment process on multiple
platforms. {\sl INDIGO--DataCloud} project contributed to open--source IaC tools such
as Ansible
and Puppet\footnote{\url{https://forge.puppet.com/indigodc}}. A representative example of
such contributions are the \textit{50 roles} developed from scratch and currently hosted in
the Ansible Galaxy portal\footnote{\url{https://galaxy.ansible.com/indigo-dc/}}. Fig.~\ref{fig:fig_confman}
shows the number of products that offer an automated means for deployment.
It shows an increase in the adoption of such tools leading to the {\sl INDIGO--1} release
(light cream points and trend line), as well as to the {\sl INDIGO--2} release
(dark blue points and trend line). The rate of adoption is lower during the weeks
before the second release because a significant fraction of the products had already adopted it
previously to the first release.
\end{itemize}

@@

\textit{Development metrics} are obtained programmatically from several sources, namely GitHub
API\footnote{\url{https://developer.github.com/v3/}.}
and the Jenkins\footnote{\url{https://jenkins.io/}} service, and graphically displayed as GitHub
pages using the GrimoireLab framework\footnote{\url{http://grimoirelab.github.io/}}.
Per--component weekly reports, including the SQA requirement
fulfilment, are issued and individually discussed through the different communication channels
with the development teams. Issue tracking metrics and KPIs are an essential piece of information
of both feature addition and defect solving, useful to detect and fix misbehaviours while in the
development phase.

@@

As any new change, involving a new feature or fix, passes through the automated
SQA machinery, the software is always ready to be released. As a result, the often compromised harmonization
between the continuous release of new features -- pushed by the development teams -- and the maintenance of
stable production systems -- demanded by users and resource providers -- can be guaranteed. The DevOps culture
theorizes and provides practical solutions that aim at unifying both software development (dev) and software
operations (ops) teams, by emphasizing the SQA techniques to avoid infrastructure disruption whenever new
developments are deployed into production systems. 
The incremental adoption of DevOps approaches throughout the project's lifetime was a key player towards
software reliability. Starting with a Continuous Integration (CI) scenario at the first stages of the project, the software delivered
was eventually validated and distributed using Continuous Delivery (CD) pipelines.

@@

Fig.~\ref{fig:fig_jenkins_CI_builds} shows the evolution, throughout the
{\sl INDIGO--DataCloud}'s first and second releases, of the required test types builds since the
implementation of the CI infrastructure. Automated functional testing coverage were not
available for all the software stack, thus the associated number of builds are fewer. 
Towards the end of the project, the software development activity slowed the pace but not completely 
ceased. The support of the CI infrastructure, once reached the project's end of life, allowed the 
continued maintenance and development of the prevailing software. Some development teams deployed 
parallel CI systems, taking advantage of the experience gained during the project.

\subsubsection{Continuous Delivery}
As DevOps suggests, frequent releases positively affect the reliability of the software as
they reduce the time in which a given defect is exposed or has an impact, allowing the 
development teams to act promptly based on the regular feedback \cite{chen2015}. The software updates of
{\sl INDIGO--DataCloud} products, which have been taking place since the second major release, are passing through
a Continuous Delivery (CD) pipeline that adds the packaging of the software right after the
successful execution of the CI pipeline previously described. Consequently, CI--validated
software components are automatically delivered in online repositories. Nonetheless,
expert supervision to validate the results is required.

@@

Two applications coming from the supported research communities,
DisVis \cite{disvis} and PowerFit \cite{powerfit}, were
integrated into a similar CI/CD pipeline described in section \ref{sec:devops}.
Fig.~\ref{fig:fig_disvis} shows the pipeline for the DisVis application.

@@

The novelty introduced in the pipeline above is the validation of the application.
Once the application is packaged as a Docker image, and subsequently uploaded
to the DockerHub repository, it is instantiated in a new container to be validated.
The application is then executed and the results compared with a set of reference outputs.
This pipeline implementation goes a step forward by testing the application
execution for the latest available Docker image in the catalogue.

@@

Two pilot infrastructures are at the disposal of developers and scientific
communities involved in the project. The aim of these testbeds is to test the
level of integration between the components involved in the {\sl INDIGO--DataCloud}
solutions and use cases validation. Within these testbeds, the applications are deployed
and executed
with the last stable version of the software in environments as similar as
possible to production ones. A map of the Pilot Preview
infrastructure is depicted in Fig.~\ref{fig:fig_pilotpreview}. It shows the
resource providers and the components or services they deployed and supported.

@@

% Comment
%%P8 C1 L26: What is the demonstration that reliability has been improved significantly with the addition of stricter methologies? Is there some quantitative evidence of uptime, or bug reports per user, or similar? Or some evidence of speed of defect detection & fixing, which you cite as greatly improved? I am very sympathetic to the idea that systematic development techniques can help improve software quality (without getting into circular definitions, where "quality" means that methodologies or standards have been applied!) but there will also be a point at which the methodology gets too heavy and compromises the "agility" of the development team: how can you identify which processes are helping and which are not?
Reliability of the software has been substantially improved by applying the SQA
criteria in the development stage. Consequently, each change in the source code
that is meant to be integrated in the production version, is validated according
to the SQA testing requirements. Thus, defects are detected and corrected early in the
software lifecycle, reducing the cost of fixing errors at later stages. Automation is
key in this process, allowing the implementation of a CI scenario that enforces
the SQA criteria compliance. The application of more advanced DevOps practices
resulted in the implementation of CD pipelines to extend automation to the
delivery of packages. Thereby, safe and frequent software releases are possible, reducing
the impact on production systems. In this regard, a pre--production validation,
through the preview testbeds or resource center's early adoption -- handled by the
Staged--rollout process -- alleviates the likelihood of discovering uncovered
issues at production.

The software lifecycle demands as well a continuous measurement and analysis
process, which allows timely reactions to discover and fix defects throughout the
development, release and maintenance stages. Conscientious metric
tracking and response sits at the core of an appropriate software maturity, as
ultimate reliability can only be achieved by enforcing and empowering the SQA
criteria through each stage of the software lifecycle.

@@

\bibitem{datagrid}
L. Momtahan and A. Martin,
``e--Science Experiences: Software Engineering Practice and the EU DataGrid,''
in \emph{Proc. 9th Asia--Pacific Softw. Eng. Conf.},
pp. 269--275, Dec. 2002.

@@

\bibitem{cordis:egi-inspire}
I. C. Plasencia,
``EGI.eu the European grid initiative,''
in \emph{Proc. 4th Iberian Grid Infra. Conf.},
pp. 5--15, 2010.

@@

\bibitem{rafi2012}
D. M. Rafi \emph{el al.},
``Benefits and limitations of automated software testing: Systematic literature review and practitioner survey,''
in \emph{Proc. 7th Int. Workshop Automation Softw. Test},
36--42, 2012.
